Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-12-16 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Resolve TODOs in manuscript

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: |
      | Methodology and Statistics
      | Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Muñoz"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: Julius Center for Health Sciences and Primary Care
    address: |
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands   
  - name: Gerko Vink
    affiliation: |
      | Methodology and Statistics
      | Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
    affiliation2: |
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, Amsterdam, The Netherlands

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. See also all of the TODOs that need to be worked out. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables"). [Just adding some text to get a better guess of what the actura abstract will look like: Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.]
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

<!-- Missing data pattern with clustering -->
<!-- TODO: simplify argument in md pattern plot and/or network plot for md patterns (summarized per variable?) -->
<!-- TODO: test if there's a random distribution of missingness in the md pattern? -->
<!-- TODO: check spss missing data pie chart? -->
<!-- TODO: similarity of patterns over clusters -->
<!-- TODO: gradient (alpha) to show in how many of the clusters the row in the pattern occurs -->
<!-- TODO: 3 columns could be 3 boxplots, over these clusters we have a distribution of these statistics across clusters.  -->
<!-- TODO: look at vim -->
<!-- TODO: check if shepley plot is an option -->

@@

Many datasets include individuals that are clustered together, for example in geographic regions, or even different studies. In the simplest case, individuals (e.g., students) are nested within a single cluster (e.g., school classes). More complex clustered structures may occur when there are multiple hierarchical levels (e.g., students in different schools or patients within hospitals within regions across countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases). With clustered data we generally assume that individuals from the same cluster tend to be more similar than individuals from other clusters. In statistical terms, this implies that observations from the same cluster are not independent and may in fact be correlated. If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01]. [TODO: make a link to imputation methods, which require adequate handling and propagation of variance; we are not recommending the adoption of multilevel models for data analysis here, but rather for imputation. VMTdJ: "I would only make that link after introducing missing data, and missing data is not mentioned yet, so I would do that in the next section"] Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations. These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'. Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.

@@

| **Concept**         | **Details**                                                                   |
|---------------------|-------------------------------------------------------------------------------|
| Sample unit         | Units of the population from which measurements are taken in a sample.        |
| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
|                     | the values observed in the study are representative of all values in the      |
|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
|                     | admission we might select only certain hospitals that are representative of   |
|                     | the difference of y between individual hospitals, but rather the variation of |
| ICC                 | The variability due to clustering is often measured by means of the           |
|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
|                     | of variance that can be attributed to the cluster-level, where a high         |
|                     | ICC would indicate that a lot of variability is due to the cluster            |
|                     | structure.                                                                    |
| Random effect       | Multilevel models typically accommodate for variability by including          |
|                     | a separate group mean for each cluster. In addition to random                 |
|                     | intercepts, multilevel models can also include random coefficients            |
|                     | and heterogeneous residual error variances across clusters [see e.g.          |
|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratification.]                     |

@@

<!-- Maybe besides CC and imputation name also likelihood based methods -->
As with any other dataset, clustered datasets may be impacted by missingness in much the same way. Several strategies can be used to handle missing data, including complete case analysis and imputation. We focus on the latter approach and discuss statistical methods for replacing the missing data with one or more plausible values. Imputation seperates the missing data problem from the analysis and the completed data can be analyzed as if it were completely observed. It is generally recommended to impute the missing values more than once to preserve uncertainty due to missingness and to allow for valid inferences (c.f. Rubin 1976).

With incomplete clustered datasets we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13]. Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18]. For example, it is possible that test results are missing for several students in one or more classes. [TODO: Provide an example for one of the case studies below.] When all observations are missing within one or more clusters, data are said to be systematically missing. [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"]

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  # coord_cartesian(expand = c(0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```
<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

@@

| **Concept**    | **Details**                                                                |
|----------------|----------------------------------------------------------------------------|
| MCAR           | Missing Completely At Random, where the probability to be missing is equal |
|                | across all data entries                                                    |
| MAR            | Missing At Random, where the probability to be missing depends on observed |
|                | information                                                                |
| MNAR           | Missing Not At Random (MNAR), where the probability to be missing          |
|                | depends on unrecorded information, making the missingness non-ignorable    |
|                | [@rubi76; @meng94].                                                        |
|                | [TODO: add congeniality, but maybe in-text?]                               |

@@

[GV: DELETE THIS SECTION?]However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent whereas multilevel data are dependent.  For this reason, we discuss how the \proglang{R} package \pkg{mice} can be used to impute multilevel data.

[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.]

@@

We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms. For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial. [TODO: change this sentence because we do not assume familiarity: "Assumed knowledge includes basic familiarity with multilevel imputation [see e.g. @audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod})"].

@@

| **Formula \pkg{lme4}**         | **Details**                                 |
|-------------------------------------------------------------|--------------------------------------------------|
| `y ~ x1 + (1 | g1)`                                      | Fixed `x1` predictor with random intercept         |
|                                                             | varying among `g1`                                  |
| `y~x1*x2+ (1| g1)`                                       | Interactions of `x1` and `x2` only in fixed effect   |
| `y ~ x1*x2+ (x2| g1)`                                      | Interactions of `x1` and `x2` only in fixed effect   |
|                                                             | with slope of `x2` randomly varying among `g1`       |
| `y ~ x1*x2+ (x1*x2| g1)`                                   | variance-covariance matrix estimated only with   |
|                                                             | the variance terms of intercept, slope of `x1`,    |
|                                                             | slope of `x2` and interaction `x1*x2`                |
| `y ~ x1*x2+ (x1 | g1)+ (x2| g1)`                            | variance-covariance matrix estimated separately, |
|                                                             | i.e, one for intercept and `x1` and another for    |
|                                                             | intercept and `x2`                                 |
| `y ~ x1 + (x1 | g1) or 1 + x1 + (1 + x1 | g1)`             | Fixed `x1` with correlated random intercept and    |
|                                                             | random slope of `x`                                |
| `y ~ x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1)` | Fixed `x1` with uncorrelated random intercept      |
|                                                             | and random slope of `x1`                           |
| `y ~ (1 | g1) + (1 | g2)`                                  | Random intercept varying among `g1` and among `g2   |
| `y ~ (1 | g1/g2) or ~ (1 | g1) + (1 | g1:g2)`                     | Random intercept varying among `g1` and `g2`         |
|                                                             | withing `g1`                                       |
|                                                             | [TODO: explain \pkg{lme4} notation here          |

@@

This tutorial is dedicated to readers who are unfamiliar with multiple imputation. More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study IMPACT) or under MNAR conditions (case study HIV).

@@

TODO: explicit statement about not going into workings of the methods. Galimer 2l methods.

## Setup
* GV: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB?

Install non-CRAN packages:
```{r packages}
devtools::install_github("amices/ggmice")
devtools::install_github("hanneoberman/miceheckman")
```

[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)        # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(miceheckman)  # for imputation cf heckman models
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

[TODO: explain case study]

@@

```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
plot_pattern(popmis)
```

@@

[TODO: Add that it doesn't work with systematic missingness (only with sporadic). There's some pros and cons, and it may not even differ much if the number of clusters is low.]

@@

[TODO: check if there is systematic missingness in this dataset, if not make Marshall Computerized Tomography classification (ct) systematically missing.]

We illustrate how to impute incomplete multilevel data by means of a case study: `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ units across $N = 15$ clusters, @metamisc]. [TODO: add more info about the complete data.] The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:

  - `name` Name of the study,
  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
  - `age` Age of the patient,
  - `motor_score` Glasgow Coma Scale motor score,
  - `pupil` Pupillary reactivity,
  - `ct` Marshall Computerized Tomography classification, [TODO: make this one var? also shows that you don't always need random effects everywhere?]
  - `hypox` Hypoxia (0=no, 1=yes),
  - `hypots` Hypotension (0=no, 1=yes),
  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  - `edh` Epidural hematoma (0=no, 1=yes),
  - `mort` 6-month mortality (0=alive, 1=dead).

@@

[TODO: show how much variance there is after different methods]

[TODO: add ICC before/after imputation and interpret: This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.]

[TODO: add descriptive statistics of the complete and incomplete data.]

@@

To explore the missingness, it is wise to look at the missing data pattern:
```{r pattern, fig.height=7.1}
plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
```

This shows... [TODO: fill in that we need to impute `ct` and `pupil`.]

To develop the best imputation model, we need to investigate the relations between the observed values of the incomplete variables and the observed values of other variables, and the relation between the missingness indicators of the incomplete variables and the observed values of the other variables. To see whether the missingness depends on the observed values of other variables, we... [TODO: fill in that we can test this statistically or use visual inspection (e.g. a histogram faceted by the missingness indicator).]

@@

## Complete case analysis [TODO: remove this?]

@@

As we can see... [TODO: fill in: This means that a higher ct (Marshall Computerized Tomography classification) is associated with a lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI ... to ..., when controlling for ...]

@@

Create a methods vector and predictor matrix, and make sure `name` is not included as predictor, but as clustering variable:
```{r impact}
meth <- make.method(dat) # methods vector
pred <- quickpred(dat)   # predictor matrix
plot_pred(pred)

pred[pred == 1] <- 2
pred["mort", ] <- 2
pred[, "mort"] <- 2
pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
pred[, "name"] <- -2
diag(pred) <- 0
plot_pred(pred)

@@

<!-- We'll now use `school` as a predictor to impute all other variables. This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16]. But at least, it includes some multilevel aspect. This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16]. Colloquially, this is 'multilevel imputation for dummies'.  -->

<!-- Add: doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->

<!-- The more the random effects are of interest, the more you need ml models. -->

@@

The imputation of these data is based on the toy example from [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R) [TODO: add reference Arxiv paper Johanna].

@@

- JOMO in \pkg{mice} -> on the side for now
<!-- # look at jomo for categorical variables? -->
<!-- # semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach -->
<!-- # pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters -->
<!-- # make assumptions of these methods explicit! -->

@@

# Think about

- Adding evaluations of the imputations such as convergence checks

- Adding some kind of help function to mice that suggests a suitable predictor matrix to the user, given a certain analysis model.

- Adding a `multilevel_ampute()` wrapper function in mice.

- Exporting `mids` objects to other packages like `lme4` or `coxme`?

- Adding a ICC=0 dataset to show that even if there is no clustering it doesn't hurt.

- Show use case for deductive imputation for cluster level variables?

- env dump in repo

>>>>

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: |
      | Methodology and Statistics
      | Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Muñoz"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
  - name: Gerko Vink
    affiliation: |
      | Methodology and Statistics
      | Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
    affiliation2: |
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, Amsterdam, The Netherlands

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables"). 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

@@

Many datasets include individuals that are clustered together, for example in geographic regions, or even different studies. In the simplest case, individuals (e.g., students) are nested within a single cluster (e.g., school classes). More complex clustered structures may occur when there are multiple hierarchical levels (e.g., students in different schools or patients within hospitals within regions across countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases). With clustered data we generally assume that individuals from the same cluster tend to be more similar than individuals from other clusters. In statistical terms, this implies that observations from the same cluster are not independent and may in fact be correlated. If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01]. Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations. These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'. Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.

@@

| **Concept**         | **Details**                                                                   |
|---------------------|-------------------------------------------------------------------------------|
| Sample unit         | Units of the population from which measurements are taken in a sample.        |
| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
|                     | the values observed in the study are representative of all values in the      |
|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
|                     | admission we might select only certain hospitals that are representative of   |
|                     | the difference of y between individual hospitals, but rather the variation of |
| ICC                 | The variability due to clustering is often measured by means of the           |
|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
|                     | of variance that can be attributed to the cluster-level, where a high         |
|                     | ICC would indicate that a lot of variability is due to the cluster            |
|                     | structure.                                                                    |
| Random effect       | Multilevel models typically accommodate for variability by including          |
|                     | a separate group mean for each cluster. In addition to random                 |
|                     | intercepts, multilevel models can also include random coefficients            |
|                     | and heterogeneous residual error variances across clusters [see e.g.          |
|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |

@@

As with any other dataset, clustered datasets may be impacted by missingness in much the same way. Several strategies can be used to handle missing data, including complete case analysis and imputation. We focus on the latter approach and discuss statistical methods for replacing the missing data with one or more plausible values. Imputation separates the missing data problem from the analysis and the completed data can be analyzed as if it were completely observed. It is generally recommended to impute the missing values more than once to preserve uncertainty due to missingness and to allow for valid inferences (c.f. Rubin 1976).

With incomplete clustered datasets we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13]. Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18]. For example, it is possible that test results are missing for several students in one or more classes. When all observations are missing within one or more clusters, data are said to be systematically missing. 
<!-- [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"] -->
<!-- TODO: Provide an example for one of the case studies below.] -->
<!-- TODO: add that indicator method doesn't work with syst missing (only sporadically). There's some pro's and con's. May not differ much if the number of clusters is low. -->

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Sporadic missingness in multilevel data", echo=FALSE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  # coord_cartesian(expand = c(0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```
<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

@@

| **Concept**    | **Details**                                                                |
|----------------|----------------------------------------------------------------------------|
| MCAR           | Missing Completely At Random, where the probability to be missing is equal |
|                | across all data entries                                                    |
| MAR            | Missing At Random, where the probability to be missing depends on observed |
|                | information                                                                |
| MNAR           | Missing Not At Random (MNAR), where the probability to be missing          |
|                | depends on unrecorded information, making the missingness non-ignorable    |
|                | [@rubi76; @meng94].                                                        |
|                |                              |

<!-- [TODO: add paragraph about congeniality and smc-fcs!]  -->

@@

<!-- [TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] -->
<!-- [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.] -->

@@

We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms. For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages [see e.g. @audi18 and @grun18] are outside the scope of this tutorial. Assumed knowledge includes basic familiarity with the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).

<!-- TODO: mention other packages? \pkg{jomo} and \pkg{mdmb} -->

@@

| **Formula \pkg{lme4}**         | **Details**                                 |
|-------------------------------------------------------------|--------------------------------------------------|
| `y ~ x1 + (1 | g1)`                                      | Fixed `x1` predictor with random intercept         |
|                                                             | varying among `g1`                                  |
| `y~x1*x2+ (1| g1)`                                       | Interactions of `x1` and `x2` only in fixed effect   |
| `y ~ x1*x2+ (x2| g1)`                                      | Interactions of `x1` and `x2` only in fixed effect   |
|                                                             | with slope of `x2` randomly varying among `g1`       |
| `y ~ x1*x2+ (x1*x2| g1)`                                   | variance-covariance matrix estimated only with   |
|                                                             | the variance terms of intercept, slope of `x1`,    |
|                                                             | slope of `x2` and interaction `x1*x2`                |
| `y ~ x1*x2+ (x1 | g1)+ (x2| g1)`                            | variance-covariance matrix estimated separately, |
|                                                             | i.e, one for intercept and `x1` and another for    |
|                                                             | intercept and `x2`                                 |
| `y ~ x1 + (x1 | g1) or 1 + x1 + (1 + x1 | g1)`             | Fixed `x1` with correlated random intercept and    |
|                                                             | random slope of `x`                                |
| `y ~ x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1)` | Fixed `x1` with uncorrelated random intercept      |
|                                                             | and random slope of `x1`                           |
| `y ~ (1 | g1) + (1 | g2)`                                  | Random intercept varying among `g1` and among `g2   |
| `y ~ (1 | g1/g2) or ~ (1 | g1) + (1 | g1:g2)`                     | Random intercept varying among `g1` and `g2`         |
|                                                             | withing `g1`                                       |
|                                                             |          |

@@

This tutorial is dedicated to readers who are unfamiliar with multiple imputation. More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study 2) or under MNAR conditions (case study 3).

@@

<!-- TODO: explicit statement about not going into workings of the methods. Galimer 2l methods. -->

## Setup
<!-- * TODO: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB? -->

Install non-CRAN packages if necessary:
```{r packages, eval=FALSE, include=TRUE}
devtools::install_github("amices/ggmice")
devtools::install_github("hanneoberman/miceheckman")
```

<!-- [TODO: Add environment info, seed and version number(s) somewhere.]  -->

Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)        # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(miceheckman)  # for imputation cf heckman models
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

<!-- [TODO: explain case study] -->

@@

```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
plot_pattern(popmis, square = TRUE)
```

@@

<!-- [TODO: Add that it doesn't work with systematic missingness (only with sporadic). There's some pros and cons, and it may not even differ much if the number of clusters is low.] -->

@@

<!-- [TODO: check if there is systematic missingness in this dataset, if not make Marshall Computerized Tomography classification (ct) systematically missing.] -->

We illustrate how to impute incomplete multilevel data by means of a case study: `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ units across $N = 15$ clusters, @metamisc]. 
<!-- [TODO: add more info about the complete data.]  -->
The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:

  - `name` Name of the study,
  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
  - `age` Age of the patient,
  - `motor_score` Glasgow Coma Scale motor score,
  - `pupil` Pupillary reactivity,
  - `ct` Marshall Computerized Tomography classification, 
  - `hypox` Hypoxia (0=no, 1=yes),
  - `hypots` Hypotension (0=no, 1=yes),
  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  - `edh` Epidural hematoma (0=no, 1=yes),
  - `mort` 6-month mortality (0=alive, 1=dead).

<!-- [TODO: make ct levels one var? also shows that you don't always need random effects everywhere?] -->

@@

<!-- [TODO: show how much variance there is after different methods] -->

<!-- [TODO: add ICC before/after imputation and interpret: This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.] -->

<!-- [TODO: add descriptive statistics of the complete and incomplete data.] -->

@@

To explore the missingness, it is wise to look at the missing data pattern. The ten most frequent missingness patterns are shown:
```{r pattern, fig.height=7.1}
plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
```

This shows that we need to impute `ct` and `pupil`.

To develop the best imputation model, we need to investigate the relations between the observed values of the incomplete variables and the observed values of other variables, and the relation between the missingness indicators of the incomplete variables and the observed values of the other variables. To see whether the missingness depends on the observed values of other variables, we can test this statistically or use visual inspection (e.g. a histogram faceted by the missingness indicator).

@@

## Complete case analysis

@@

As we can see, a higher `ct` (Marshall Computerized Tomography classification) is associated with a lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI ... to ..., when controlling for...

<!-- TODO: fill in results above -->

@@

Create a methods vector and predictor matrix, and make sure `name` is not included as predictor, but as clustering variable:
```{r impact}
meth <- make.method(dat) # methods vector
pred <- quickpred(dat)   # predictor matrix
plot_pred(pred, rotate = TRUE)

pred[pred == 1] <- 2
pred["mort", ] <- 2
pred[, "mort"] <- 2
pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
pred[, "name"] <- -2
diag(pred) <- 0
plot_pred(pred, rotate = TRUE)

@@

@@

The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)

<!-- [TODO: add reference Arxiv paper Johanna].  -->

@@

<!-- - JOMO in \pkg{mice} -> on the side for now -->
<!-- # look at jomo for categorical variables? -->
<!-- # semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach -->
<!-- # pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters -->
<!-- # make assumptions of these methods explicit! -->

@@

<!-- # Think about -->

<!-- - Adding evaluations of the imputations such as convergence checks -->

<!-- TODO in the future: Adding some kind of help function to mice that suggests a suitable predictor matrix to the user, given a certain analysis model. -->

<!-- TODO in the future: Adding a `multilevel_ampute()` wrapper function in mice. -->

<!-- - Exporting `mids` objects to other packages like `lme4` or `coxme`?  -->

<!-- - Adding a ICC=0 dataset to show that even if there is no clustering it doesn't hurt. -->

<!-- - Show use case for deductive imputation for cluster level variables? -->

<!-- - env dump in repo -->
